---
title: "Cookies and analytics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cookies and analytics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
library(shinyGovstyle)
```

## Overview

Public services need to tell users about cookies and, where they set
non-essential cookies (such as analytics), let users accept or reject them.
shinyGovstyle gives you the building blocks for this:

- `cookieBanner()`: the GOV.UK styled banner that appears at the top of the
  page, with accept / reject / hide message states wired up for you.
- `radio_button_Input()`: a GOV.UK styled radio group, useful for a "change
  your cookie settings" control on a dedicated cookies page.
- `update_radio_button_Input()`: the server-side companion that lets you
  change the selected option (or the choices, or the label) of a radio group
  from your server code.

shinyGovstyle is deliberately **agnostic about analytics**. It does not set,
read, or send any analytics cookies itself. It gives you the consent UI and the
reactive plumbing; you decide what to do with the user's choice. The
[Extending for analytics](#extending-for-analytics) section points to a worked
example from DfE using Google Analytics.

## The cookie banner

`cookieBanner()` needs `shinyjs::useShinyjs()` to be present so the banner can
show and hide its different states. All of the ids are preset, so the server
logic is a fixed set of `observeEvent()` handlers.

```{r}
library(shiny)
library(shinyGovstyle)

ui <- fluidPage(
  shinyjs::useShinyjs(),
  shinyGovstyle::header(
    org_name = "Example",
    service_name = "User Examples"
  ),
  shinyGovstyle::cookieBanner("User Examples"),
  shinyGovstyle::gov_layout(size = "two-thirds"),
  shinyGovstyle::footer(full = TRUE)
)

server <- function(input, output, session) {
  # Show the relevant confirmation message and hide the main banner
  observeEvent(input$cookieAccept, {
    shinyjs::show(id = "cookieAcceptDiv")
    shinyjs::hide(id = "cookieMain")
  })

  observeEvent(input$cookieReject, {
    shinyjs::show(id = "cookieRejectDiv")
    shinyjs::hide(id = "cookieMain")
  })

  # Hide the whole banner once the user has read the confirmation
  observeEvent(input$hideAccept, shinyjs::toggle(id = "cookieDiv"))
  observeEvent(input$hideReject, shinyjs::toggle(id = "cookieDiv"))

  # input$cookieLink can take users to a dedicated cookies page, for example
  # via shiny::updateTabsetPanel()
  observeEvent(input$cookieLink, {
    # navigate to your cookies page here
  })
}

if (interactive()) shinyApp(ui, server)
```

## A cookies settings page

A common pattern is a dedicated cookies page with a radio group that lets the
user change their analytics choice. Build the control with
`radio_button_Input()`:

```{r}
shinyGovstyle::radio_button_Input(
  inputId = "cookies_analytics",
  label = "Do you want to accept analytics cookies?",
  choices = c("Yes" = "yes", "No" = "no"),
  selected = "no",
  inline = TRUE
)
```

The challenge is keeping this radio in sync with the choice the user already
made in the banner. When the user clicks "Accept analytics cookies", the radio
on the settings page should reflect that without the user touching it. That is
what `update_radio_button_Input()` is for:

```{r}
server <- function(input, output, session) {
  observeEvent(input$cookieAccept, {
    shinyGovstyle::update_radio_button_Input(
      session,
      inputId = "cookies_analytics",
      selected = "yes"
    )
  })

  observeEvent(input$cookieReject, {
    shinyGovstyle::update_radio_button_Input(
      session,
      inputId = "cookies_analytics",
      selected = "no"
    )
  })
}
```

`update_radio_button_Input()` mirrors `shiny::updateRadioButtons()`: only the
arguments you pass are sent to the client, and the `inputId` is namespaced
automatically when you call it inside a Shiny module. That means you can use
the GOV.UK styled radio everywhere, including inside modules, and still drive
its value programmatically.

## Extending for analytics

shinyGovstyle stops at the consent UI on purpose. To actually act on the
user's choice (for example to load Google Analytics only after consent, store
the decision in a cookie, and respond to it across sessions), you layer your
own logic (or another package's) on top of these components.

[dfeshiny](https://github.com/dfe-analytical-services/dfeshiny) is a worked
example of exactly this. It provides a cookies module
(`cookies_banner_ui()` / `cookies_banner_server()` and `cookies_panel_ui()`)
that builds on shinyGovstyle's components and adds Google Analytics consent
handling: reading and writing the consent cookie, and toggling analytics on or
off based on the user's choice. If you need a ready-made analytics consent
flow rather than the building blocks, that module is a good place to start,
and a good template if you are writing your own.
