---
title: "Layout options"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Layout options}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(shinyGovstyle)
```

## Overview

shinyGovstyle provides a set of layout functions that produce the HTML structure GOV.UK Frontend CSS expects. This vignette covers all of the layout functions available and explains how they fit together to build a complete app.

The layout functions fall into two groups:

- **Page-level components** — `header()`, `footer()`, `banner()`, `cookieBanner()`, `skip_to_main()`, and `service_navigation()` — form the frame of the page that sits outside the main content area.
- **Content layout functions** — `gov_main_layout()`, `gov_row()`, `gov_box()`, and `gov_layout()` — structure content within the main content area.

---

## Page-level components

These components form the outer frame of every page. They sit outside the main content area and are consistent across all pages of your app.

```
+-------------------------------------------------------+
|  skip_to_main()   [visually hidden, keyboard only]    |
+-------------------------------------------------------+
|  cookieBanner()   [optional]                          |
+-------------------------------------------------------+
|  header()                                             |
+-------------------------------------------------------+
|  service_navigation()   [optional, multi-page apps]   |
+-------------------------------------------------------+
|  banner()   [optional, e.g. Beta or Alpha]            |
+-------------------------------------------------------+
|                                                       |
|  gov_main_layout()   ← id = "main"                    |
|  +--------------------------------------------------+  |
|  |  your content goes here                         |  |
|  +--------------------------------------------------+  |
|                                                       |
+-------------------------------------------------------+
|  footer()                                             |
+-------------------------------------------------------+
```

### `skip_to_main()`

Provides a visually hidden "Skip to main content" link that becomes visible when focused by a keyboard user. This is an accessibility requirement and should always be the first element in your UI, before the header.

```{r, eval = FALSE}
skip_to_main()
```

By default it links to `#main`, which matches the `id` applied by `gov_main_layout()`. If you change the `inputID` argument of `gov_main_layout()`, pass the same value to `skip_to_main()`.

For more information, read the documentation for the [GOV.UK Skip link component](https://design-system.service.gov.uk/components/skip-link/).

### `cookieBanner()`

Displays a GOV.UK-styled cookie consent banner. It requires `shinyjs::useShinyjs()` to be present in the UI. All element IDs within the banner are preset — see `?cookieBanner` for the server-side `observeEvent` pattern needed to handle accept and reject interactions.

```{r, eval = FALSE}
shinyjs::useShinyjs()
cookieBanner("My service name")
```

For more information, including when this should be used, read the documentation for the [GOV.UK Cookie banner component](https://design-system.service.gov.uk/components/cookie-banner/).

### `header()`

Creates a GOV.UK styled header bar, optionally containing your department logo, name and service name. This is not the official [GOV.UK header](https://design-system.service.gov.uk/components/header/), as that should only be used on GOV.UK domains. If you believe you have an R Shiny app on a GOV.UK domain, please [raise an issue](https://github.com/dfe-analytical-services/shinyGovstyle/issues/new/choose)
to request this as an addition to the package.

```{r, eval = FALSE}
header(
  org_name = "Department for Education",
  service_name = "My dashboard"
)
```

### `banner()`

Displays a phase banner immediately below the header, used to indicate the maturity of your service and give a clear route for users to provide feedback.

```{r, eval = FALSE}
banner(
  inputId = "phase-banner",
  type = "Beta",
  label = paste0(
    "This is a new service \u2014 your ",
    "<a class="govuk-link" href="#">feedback</a> will help us to improve it."
  )
)
```

For more information on when and how to use this, read the documentation for the [GOV.UK Phase banner component](https://design-system.service.gov.uk/components/phase-banner/). 


### `footer()`

Creates a GOV.UK styled footer, though like the header, this is not an offical version as that should only be used on a GOV.UK domain. Use `full = TRUE` to include the OGL licence logo and Crown copyright statement. You can add support links that point either to internal hidden tab panels or to external URLs.

```{r, eval = FALSE}
# Minimal footer
footer()

# Footer with support links
footer(
  links = c(
    `Accessibility statement` = "accessibility_footer_link",
    `Cookies` = "cookies_footer_link"
  )
)
```

Internal links use auto-generated inputIDs — the link text lowercased with non-alphanumeric characters replaced by underscores — that you handle with `observeEvent()` in your server to switch the active tab panel.

---

## The main content area

`gov_main_layout()` produces a `<div class="govuk-width-container">` wrapping a `<main class="govuk-main-wrapper">`. The outer `<div>` constrains content width; the `<main>` element carries the responsive vertical padding. Everything between the page-level components and the footer lives inside it.

```{r, eval = FALSE}
gov_main_layout(
  # your content here
)
```

The `id` (default `"main"`) is applied directly to the `<main>` element, which is the correct target for `skip_to_main()`. The `<main>` element also carries `role="main"` and `tabindex="-1"`, so keyboard focus moves to it when the skip link is activated.

---

## The primary layout system

Inside `gov_main_layout()`, content is structured using a three-function grid system: `gov_row()`, `gov_box()`, and optionally `gov_text()`.

```
gov_main_layout()
└── gov_row()
    ├── gov_box(size = "two-thirds")
    │   └── [your content]
    └── gov_box(size = "one-third")
        └── [your content]
```

### `gov_row()`

Creates a GOV.UK grid row. You can have multiple rows inside `gov_main_layout()`, each stacked vertically.

```{r, eval = FALSE}
gov_main_layout(
  gov_row(
    # columns go here
  ),
  gov_row(
    # another row
  )
)
```

### `gov_box()`

Creates a column within a row. The `size` argument controls the column width using GOV.UK Frontend's grid classes:

| `size` | Width |
|---|---|
| `"full"` | 100% |
| `"one-half"` | 50% |
| `"two-thirds"` | 66% |
| `"one-third"` | 33% |
| `"three-quarters"` | 75% |
| `"one-quarter"` | 25% |

Sizes within a row should add up to a full width. For example, `"two-thirds"` and `"one-third"` sit side by side:

```{r, eval = FALSE}
gov_main_layout(
  gov_row(
    gov_box(
      size = "two-thirds",
      heading_text("Main content", size = "l"),
      # inputs, text, etc.
    ),
    gov_box(
      size = "one-third",
      heading_text("Sidebar", size = "m"),
      # supporting content
    )
  )
)
```

For a simple single-column layout, use `size = "full"`:

```{r, eval = FALSE}
gov_main_layout(
  gov_row(
    gov_box(
      size = "full",
      heading_text("Page title", size = "l")
    )
  )
)
```

### `gov_text()`

A wrapper that produces a `<p class="govuk-body">` paragraph element. For full guidance on `gov_text()` and all other text functions, see the [Headings and text](headings-and-text.html) vignette.

---

## `gov_layout()` — legacy alternative

> **Warning:** `gov_layout()` is not recommended for new development and may be removed in a future release. Use `gov_main_layout()` with `gov_row()` and `gov_box()` instead.

`gov_layout()` is a single-function alternative that combines a width container and a column in one call:

```{r, eval = FALSE}
gov_layout(
  size = "two-thirds",
  heading_text("Page title", size = "l"),
  # content
)
```

It is well suited to simple, single-column apps where you want a width constraint without setting up the full `gov_main_layout()` / `gov_row()` / `gov_box()` hierarchy.

**As soon as your app needs more than one column, multiple rows, or a combination of widths, switch to the full system.** Nesting `gov_layout()` inside `gov_main_layout()` will produce doubled-up width container HTML and cause the content to appear visually inset from the page-level components.

---

## Multi-page dashboards

For apps with multiple sections, use `service_navigation()` in combination with a hidden tab panel. The navigation bar renders as a row of links below the header; clicking a link fires a Shiny input that you use in your server to switch the visible panel.

### Setting up navigation links

Pass a named character vector to `service_navigation()`. The names are displayed as link text; the values become the inputIDs:

```{r, eval = FALSE}
service_navigation(
  c(
    "Summary" = "nav_summary",
    "Detailed data" = "nav_detail",
    "User guide" = "nav_guide"
  )
)
```

If you pass an unnamed vector, inputIDs are auto-generated by lowercasing the text and replacing non-alphanumeric characters with underscores (e.g. `"Detailed data"` becomes `detailed_data`).

### Wiring navigation to panels

Use a hidden tab panel for the content area and `observeEvent()` in your server to switch panels when a navigation link is clicked. When the user clicks a service navigation link, the JavaScript binding updates the active state automatically — you only need to switch the panel:

```{r, eval = FALSE}
# ui.R — shiny tabsetPanel
shiny::tabsetPanel(
  type = "hidden",
  id = "main_panels",
  shiny::tabPanel("Summary",       value = "nav_summary",  "Content"),
  shiny::tabPanel("Detailed data", value = "nav_detail",   "Content"),
  shiny::tabPanel("User guide",    value = "nav_guide",    "Content")
)

# server.R — nav link click: JS handles the active state, just switch the panel
shiny::observeEvent(input$nav_summary, {
  shiny::updateTabsetPanel(session, "main_panels", selected = "nav_summary")
})
```

If you prefer bslib tab panels, use `bslib::navset_hidden()` and `bslib::nav_select()` instead:

```{r, eval = FALSE}
# ui.R — bslib navset_hidden
bslib::navset_hidden(
  id = "main_panels",
  bslib::nav_panel("Summary",       value = "nav_summary",  "Content"),
  bslib::nav_panel("Detailed data", value = "nav_detail",   "Content"),
  bslib::nav_panel("User guide",    value = "nav_guide",    "Content")
)

# server.R
shiny::observeEvent(input$nav_summary, {
  bslib::nav_select("main_panels", "nav_summary")
})
```

Repeat the `observeEvent` block for each navigation link.

`update_service_navigation()` is only needed when navigation is triggered **programmatically** — for example, via a next / back button — because in that case the nav link itself is not clicked and the active state does not update automatically. See `?update_service_navigation` for full details and examples.

```{r, eval = FALSE}
# server.R — programmatic navigation: must update both the panel and the nav
shiny::observeEvent(input$next_btn, {
  shiny::updateTabsetPanel(session, "main_panels", selected = "nav_detail")
  shinyGovstyle::update_service_navigation(session, "nav_detail")
})
```

### Footer-only pages

Some pages — such as an accessibility statement, privacy notice, or cookies information page — should not appear in the service navigation but still need to be reachable. The standard pattern is to add a link in `footer()` and a corresponding hidden tab panel, but to omit the link from `service_navigation()`.

Because the user navigates to these pages outside of the service navigation, there is no active nav item to highlight. You do not need to call `update_service_navigation()` for these transitions. However, you should call it when navigating *back* to a main page from a footer-linked page, so the correct nav item becomes active again.

```{r, eval = FALSE}
# ui.R — footer link, no entry in service_navigation()
footer(
  full = TRUE,
  links = c(`Accessibility statement` = "accessibility_footer_link")
)

# ui.R — tab panel exists in the hidden tabset but not in service_navigation()
shiny::tabsetPanel(
  type = "hidden",
  id = "main_panels",
  shiny::tabPanel("Summary", value = "nav_summary", "Content"),
  shiny::tabPanel("Accessibility statement", value = "accessibility_panel",
                  "Content")
)

# server.R — navigate to the footer page (no update_service_navigation needed)
shiny::observeEvent(input$accessibility_footer_link, {
  shiny::updateTabsetPanel(session, "main_panels",
                           selected = "accessibility_panel")
})
```

### Modularising the code

Once an app has multiple pages, it is strongly recommended to use Shiny modules to keep each page's UI and server logic self-contained. The `inst/example_app` bundled with this package demonstrates this pattern: each page is a module in `inst/example_app/modules/`, with `mod_<name>_ui()` and `mod_<name>_server()` functions called from the top-level `ui.R` and `server.R`. This keeps individual files focused and makes it straightforward to add or remove pages without touching the overall app structure.

---

## Complete example

The following is a minimal but complete multi-page app that uses all of the layout components covered in this vignette:

```{r, eval = FALSE}
library(shiny)
library(shinyGovstyle)

ui <- bslib::page_fluid(
  skip_to_main(),
  header(
    org_name = "My department",
    service_name = "My dashboard"
  ),
  service_navigation(
    c(
      "Summary"  = "nav_summary",
      "About"    = "nav_about"
    )
  ),
  banner(
    inputId = "phase",
    type = "Beta",
    label = "This is a new service."
  ),

  gov_main_layout(
    shiny::tabsetPanel(
      type = "hidden",
      id = "main_panels",

      shiny::tabPanel(
        "Summary", value = "nav_summary",
        gov_row(
          gov_box(
            size = "two-thirds",
            heading_text("Summary", size = "l"),
            gov_text("Welcome to the summary page.")
          ),
          gov_box(
            size = "one-third",
            heading_text("Quick facts", size = "m"),
            gov_text("Supporting information goes here.")
          )
        )
      ),

      shiny::tabPanel(
        "About", value = "nav_about",
        gov_row(
          gov_box(
            size = "full",
            heading_text("About this dashboard", size = "l"),
            gov_text("This page describes the dashboard.")
          )
        )
      ),

      shiny::tabPanel(
        "Accessibility statement", value = "accessibility_panel",
        gov_row(
          gov_box(
            size = "full",
            heading_text("Accessibility statement", size = "l"),
            gov_text("This page describes the accessibility of the dashboard.")
          )
        )
      )
    )
  ),

  footer(
    links = c(`Accessibility statement` = "accessibility_footer_link")
  )
)

server <- function(input, output, session) {
  shiny::observeEvent(input$nav_summary, {
    shiny::updateTabsetPanel(session, "main_panels", selected = "nav_summary")
  })
  shiny::observeEvent(input$nav_about, {
    shiny::updateTabsetPanel(session, "main_panels", selected = "nav_about")
  })
  shiny::observeEvent(input$accessibility_footer_link, {
    shiny::updateTabsetPanel(session, "main_panels",
                             selected = "accessibility_panel")
  })
}

shiny::shinyApp(ui, server)
```
