---
title: "Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(shinyGovstyle)
```

## Overview
**shinyGovstyle** integrates [GOV.UK Design System](https://design-system.service.gov.uk/) components into R Shiny. It aims to make it easier for government analysts to develop production-ready dashboards that users find intuitive and can trust. 

The [GOV.UK Design System component documentation](https://design-system.service.gov.uk/components/) remains the authoritative reference for all components, including detailed guidance of when, how, and when not(!) to use each of the components.

> **Note:** Using accessible components does not replace the need for comprehensive accessibility testing of the combination you put together in your own specific application.

We welcome feedback and contributions:

- **Suggestions, requests and bugs:** [Raise a GitHub issue](https://github.com/dfe-analytical-services/shinyGovstyle/issues/new) for bugs, requests or suggestions.
- **Search:** Check [existing issues](https://github.com/dfe-analytical-services/shinyGovstyle/issues) before posting.
- **Contribute:** See our [contributing guide](https://github.com/dfe-analytical-services/shinyGovstyle/blob/main/.github/CONTRIBUTING.md) to help improve the package.

---

## See it in action

The quickest way to get a feel for what is available is the example showcase app, which demonstrates the full range of components:

- **Online:** [department-for-education.shinyapps.io/shinygovstyle-example-app](https://department-for-education.shinyapps.io/shinygovstyle-example-app/)
- **Local** (after installing the package): `shinyGovstyle::run_example()`

The source code for the example app lives in [`inst/example_app/`](https://github.com/dfe-analytical-services/shinyGovstyle/tree/main/inst/example_app) and is a useful reference for how components fit together in a real multi-page app using [Shiny modules](https://mastering-shiny.org/scaling-modules.html).

---

## Installation

Install the latest stable release from CRAN:

```{r, eval = FALSE}
install.packages("shinyGovstyle")
```

To use the latest development version directly from GitHub:

```{r, eval = FALSE}
pak::pak("dfe-analytical-services/shinyGovstyle")
```

---

## shinyjs dependency

Some components require [`shinyjs`](https://deanattali.com/shinyjs/) to be loaded in your UI. Add `shinyjs::useShinyjs()` anywhere in your UI when using:

- `error_on()` / `error_off()` — inline field-level error toggling
- `word_count()` — live character and word counting on text inputs
- `cookieBanner()` — GOV.UK cookie consent banner

```{r, eval = FALSE}
ui <- bslib::page_fluid(
  shinyjs::useShinyjs(),
  # rest of your UI
)
```

---

## A minimal working app

The skeleton below shows the components that appear on every page of a shinyGovstyle app. It uses [`bslib`](https://rstudio.github.io/bslib/), which is generally [recommended for building responsive apps with R Shiny](https://shiny.posit.co/r/getstarted/shiny-basics/lesson1/).

```{r, eval = FALSE}
library(shiny)
library(shinyGovstyle)

ui <- bslib::page_fluid(
  skip_to_main(),
  header(
    main_text = "My department",
    secondary_text = "My dashboard"
  ),
  banner(
    inputId = "phase",
    type = "Beta",
    label = "This is a new service."
  ),
  gov_main_layout(
    gov_row(
      gov_box(
        size = "two-thirds",
        heading_text("Welcome", size = "l", level = 1),
        gov_text("Select a report from the navigation above.")
      )
    )
  ),
  footer()
)

server <- function(input, output, session) {}

shiny::shinyApp(ui, server)
```

For a more complete example - including multi-page navigation, footer links, and the server logic to wire it all together - see the [complete example](layout-options.html#complete-example) at the end of the Layout options vignette.

---

## Component overview

### Page structure and layout

All page-level components (`header()`, `footer()`, `banner()`, `skip_to_main()`, `service_navigation()`, `cookieBanner()`) and the content grid system (`gov_main_layout()`, `gov_row()`, `gov_box()`) are covered in the [Layout options](layout-options.html) vignette.

### Headings and text

Styled headings, body text, lists, callouts, external links, notification banners, and typography are covered in the [Headings and text](headings-and-text.html) vignette.

### Form inputs

`text_input()`, `text_area_input()`, `select_input()`, `checkbox_input()`, `radio_button_Input()`, `date_Input()`, `file_Input()`, `button_Input()`, and `download_button()` provide GOV.UK-styled equivalents of standard Shiny form controls; values are read from `input$<inputId>` in the server in the usual way.

### Interactive components

`accordion()` and `govTabs()` provide collapsible panels and tabbed interfaces with full keyboard navigation and ARIA support built in.

### Error handling

`error_summary()` produces a GOV.UK error summary panel that lists all validation errors at the top of the page; `error_on()` and `error_off()` toggle inline field-level error messages on any input. These require `shinyjs::useShinyjs()` in the UI, see [shinyjs dependency](#shinyjs-dependency) for more details.

### Data display

`govTable()` and `govReactable()` render data tables with GOV.UK styling; `value_box()` displays a headline figure with a label, suited to summary statistics at the top of a dashboard.
