---
title: "Tables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(shinyGovstyle)
```

## Before you add a table

Use a table when users need to compare values across rows and columns. Do not use one to lay out a page. Use `gov_layout()` or `gov_row()` for that instead.

If your data is a handful of numbers, a sentence or a `value_box()` is often clearer than a table. The [GOV.UK guidance on tables](https://guidance.publishing.service.gov.uk/formatting-content/text-formatting/tables/) has more on when a table helps and how to keep one simple.

## Choosing a table function

shinyGovstyle has three ways to show a table:

- `govTable()` for small tables that do not change
- `govReactable()` for bigger tables, or when users need to sort or page through the data
- `govReactableOutput()` with `renderGovReactable()` for tables in a Shiny app that change as users pick options

`govTable()` builds a plain HTML table in the GOV.UK style. It is the simplest option and works well up to a few dozen rows. It warns you if you give it more than 50 rows, because a long static table is slow to load and hard to read.

`govReactable()` uses the [reactable](https://glin.github.io/reactable/) package, so users can sort columns and move between pages. It also works in R Markdown and Quarto documents.

## Giving a table a title

Every table needs a title, and users of screen readers hear it when they reach the table. All three functions take a `caption` for the title and an optional `subtitle` for the line underneath.

The [Analysis Function guidance on tables](https://analysisfunction.civilservice.gov.uk/policy-store/data-visualisation-tables/) suggests a title in two parts. The first is a short headline that says what the data shows. The second says what the data is, where it covers and when. Put the headline in `caption` and the rest in `subtitle`:

```{r, eval = FALSE}
govTable(
  "costs_table",
  transport_data_small,
  caption = "Bike and car costs were highest in March",
  subtitle = "Cost of bikes and cars (£), England, January to March 2025"
)
```

If every value in the table uses the same unit, put the unit in the subtitle, as with "(£)" above. If units differ, put them in the column headings.

A headline is optional. If you do not want one, put the what, where and when in `caption` and leave out `subtitle`.

Put the source of the data under the table, for example with `gov_text()`, rather than in the title.

## Small tables with govTable()

`govTable()` needs an id, a data frame and a caption.

```{r, eval = FALSE}
govTable(
  "costs_table",
  transport_data_small,
  caption = "Bike and car costs were highest in March",
  subtitle = "Cost of bikes and cars (£), England, January to March 2025",
  num_col = c(2, 3)
)
```

`num_col` right-aligns the numeric columns, given by position. GOV.UK tables right-align numbers so they are easier to compare down a column.

To set column widths, give `width_overwrite` one width for each column, such as `c("one-half", "one-quarter", "one-quarter")`.

## Larger tables with govReactable()

```{r, eval = FALSE}
govReactable(
  transport_data,
  caption = "Costs peaked in March for every vehicle type",
  subtitle = "Cost of bikes, vans and buses (£), England, January to May 2025",
  page_size = 5
)
```

Numeric columns are right-aligned for you, so numbers line up for comparison. If a column holds numbers as text, such as "£85", name it in `right_col` to right-align it too. `page_size` sets how many rows show on each page.

To change how a single column looks, for example to round numbers, pass `columns` a list of `reactable::colDef()` settings. Columns you do not mention keep the GOV.UK styling.

```{r, eval = FALSE}
govReactable(
  transport_data,
  caption = "Costs peaked in March for every vehicle type",
  columns = list(
    bikes = reactable::colDef(format = reactable::colFormat(prefix = "£"))
  )
)
```

govReactable does not show reactable's sort icons, because they are not accessible. Users can still sort by selecting a column heading.

## Tables in a Shiny app

In a Shiny app, put `govReactableOutput()` in the UI and build the table in the server with `renderGovReactable()`. The title goes on `govReactableOutput()`, because it stays on the page while the table data changes.

```{r, eval = FALSE}
ui <- gov_page(
  govReactableOutput(
    "costs_table",
    caption = "Costs peaked in March for every vehicle type",
    subtitle = "Cost of bikes, vans and buses (£), England, January to May 2025"
  )
)

server <- function(input, output, session) {
  output$costs_table <- renderGovReactable({
    govReactable(transport_data)
  })
}
```

### Changing the title when the data changes

If users can filter the table, the title should describe what they are looking at. Use `update_reactable_caption()` in the server to change the caption, the subtitle, or both.

```{r, eval = FALSE}
colours <- sort(unique(transport_data$colours))

ui <- gov_page(
  select_Input(
    inputId = "colour",
    label = "Colour",
    select_text = colours,
    select_value = colours
  ),
  govReactableOutput(
    "costs_table",
    caption = paste(colours[1], "vehicle costs peaked in March"),
    subtitle = paste(colours[1], "vehicle costs (£), January to May")
  )
)

server <- function(input, output, session) {
  output$costs_table <- renderGovReactable({
    govReactable(subset(transport_data, colours == input$colour))
  })

  shiny::observe({
    update_reactable_caption(
      session,
      "costs_table",
      caption = paste(input$colour, "vehicle costs peaked in March"),
      subtitle = paste(input$colour, "vehicle costs (£), January to May")
    )
  })
}
```

Start the caption in the UI with the text for the default choice, so the title is right before the server runs.

A subtitle can only be updated if `govReactableOutput()` was given one to start with.

The new title is not read out when it changes. Users hear it when they next reach the table. This avoids an announcement every time they change a filter.

## Further guidance

- [GOV.UK Design System table component](https://design-system.service.gov.uk/components/table/)
- [GOV.UK guidance on using tables](https://guidance.publishing.service.gov.uk/formatting-content/text-formatting/tables/)
- [Analysis Function guidance on tables](https://analysisfunction.civilservice.gov.uk/policy-store/data-visualisation-tables/)
